---
title: "Cyclistic Case Study Solution with R"
author: "Göktuğ Yaman"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setting the environment
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r, echo=FALSE}
setwd("/Users/goktugyaman/Documents/Data_Analysis_Case_Study")
all_trips_v6 <- read_csv("all_trips_v6.csv")
all_trips_v6$year <- format(as.Date(all_trips_v6$started_at), "%Y")
```

### Preparing the working directory
setwd("/Users/goktugyaman/Documents/Data_Analysis_Case_Study/CSV")

### Data collection: importing csv files into RStuido
all_2013 <- read_csv("2013.csv")
q3_2014_07 <- read_csv("2014_07.csv") ...

### Standardizing tables in order to join them into single dataframe
all_2013 <- rename(all_2013,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = starttime,
                   ended_at = stoptime,
                   start_station_id = from_station_id,
                   start_station_name = from_station_name,
                   end_station_id = to_station_id,
                   end_station_name = to_station_name,
                   member_casual = usertype) ...

### Converting data types into characters to stack them together

#### ride_id and rideable_type turned into characters
all_2013 <-  mutate(all_2013, ride_id = as.character(ride_id),
                    rideable_type = as.character(rideable_type))  ...

#### Date columns turned into POSIXct, tables with problems to convert data type have been exported into excel and problem fixed

##### Setting the environment
```{r}
library(readxl)
```

##### Setting the working directory for excel files
setwd("/Users/goktugyaman/Documents/Data_Analysis_Case_Study/Excels")

##### Importing the files after csv's splitted through terminal via split command
n_q1_q2_2014 <- read_excel("2014_Q1Q2.xlsx")
n_q3_2014_07 <- read_excel("2014_Q3_07.xlsx") ...

##### Adujusting column names of the excel files
n_q1_q2_2014 <- rename(n_q1_q2_2014,
                       ride_id = trip_id,
                       started_at = start_time,
                       ended_at = stop_time,
                       start_station_id = from_station_id,
                       start_station_name = from_station_name,
                       end_station_id = to_station_id,
                       end_station_name = to_station_name,
                       member_casual = user_type) ...

##### Converting ride_id, end_station_id, and start_staiton_id columns into character type
n_q1_q2_2014 <- mutate(n_q1_q2_2014, ride_id = as.character(ride_id)) ...

### Combining tables
all_trips <- bind_rows(all_2013,
                       n_q1_q2_2014, n_q3_2014_07, n_q3_2014_08_09, n_q4_2014,
                       n_q1_2015, n_q2_2015, n_q3_2015_07, n_q3_2015_08, n_q3_2015_09, n_q4_2015, ...

### Standardizing member_casual column, member for subscribers and casual for casual riders
all_trips <- mutate(all_trips, member_casual = recode(member_casual,
                                                      "Subscriber" = "member",
                                                      "Customer" = "casual",
                                                      "Dependent" = "casual")) ...

### Adding columns for aggregation
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

### Adding a column for ride_length into whole table
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

### Dropping inconsistent and unrelated columns to analysis
all_trips <- all_trips %>%  select(-c("start_station_id",
                                      "end_station_id",
                                      "start_station_name",
                                      "end_station_name",
                                      "birthyear",
                                      "gender",
                                      "birthday",
                                      "trip_day",
                                      "01 - Rental Details Duration In Seconds Uncapped",
                                      "Member Gender",
                                      "05 - Member Details Member Birthday Year",
                                      "start_lat",
                                      "start_lng",
                                      "end_lat",
                                      "end_lng",
                                      "tripduration",
                                      "ride_id",
                                      "rideable_type"))

### Removing the trips that lasted less than 120 seconds in order to eliminate test rides and untypical rides
all_trips_v2 <- all_trips[!(all_trips$ride_length<120),]
all_trips_v3 <- all_trips_v2

### Fixing the ride lengths decimal difference of years 2014-2017
all_trips_v3$ride_length <- ifelse(all_trips_v3$year == 2015, all_trips_v3$ride_length/1000, all_trips_v3$ride_length)
rm(all_trips_v2) ...

### Fixing the ride lengths decimal difference of years 2014-2017
all_trips_v3$ride_length <- ifelse(all_trips_v3$year == 2015, all_trips_v3$ride_length/1000, all_trips_v3$ride_length)
rm(all_trips_v2)
all_trips_v4 <- all_trips_v3
all_trips_v4$ride_length <- ifelse(all_trips_v4$year == 2016, all_trips_v4$ride_length/1000, all_trips_v4$ride_length)
rm(all_trips_v3)
all_trips_v5 <- all_trips_v4
all_trips_v5$ride_length <- ifelse(all_trips_v5$year == 2017, all_trips_v5$ride_length/1000, all_trips_v5$ride_length)
all_trips_v6 <- all_trips_v5
rm(all_trips_v4)
all_trips_v6$ride_length <- ifelse(all_trips_v6$year == 2014, all_trips_v6$ride_length/10, all_trips_v6$ride_length)
rm(all_trips_v5)

### Sorting the data according to weekday
```{r}
all_trips_v6$day_of_week <- ordered(all_trips_v6$day_of_week, 
                                    levels=c("Monday", "Tuesday", "Wednesday", "Thursday",                                                    "Friday", "Saturday", "Sunday"))
```

### Summary statistics (values in seconds)

##### Average trip duration for both subscribers and casual riders
```{r}
aggregate(all_trips_v6$ride_length ~ all_trips_v6$member_casual, FUN = mean)
```

##### Median value of trip duration for both customer groups
```{r}
aggregate(all_trips_v6$ride_length ~ all_trips_v6$member_casual, FUN = median)
```

##### Maximum trip duration for both customer groups
```{r}
aggregate(all_trips_v6$ride_length ~ all_trips_v6$member_casual, FUN = max)
```

##### Minimum trip duration for both customer groups
```{r}
aggregate(all_trips_v6$ride_length ~ all_trips_v6$member_casual, FUN = min)
```

##### Average trip duration for both customer groups, grouped by weekdays
```{r}
aggregate(all_trips_v6$ride_length ~ all_trips_v6$member_casual + all_trips_v6$day_of_week, FUN = mean)
```

### Analysis on total number of rides

#### Daily
The daily number of different customer groups shows that the number of members is always higher than the casual riders.
The number of members during weekends is lower than on weekdays but it is vice versa for the casual riders.
```{r, echo=FALSE}
all_trips_v6 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy",
                               "maroon")) +
  labs(title = "Daily Number of Trips Taken", fill = "Member or Casual", x = "", y = "Number of Daily Trips")
```

#### Monthly
```{r, echo=FALSE}
all_trips_v6 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()) %>%  
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy",
                               "maroon")) +
  labs(title = "Monthly Number of Trips Taken", fill = "Member or Casual", x = "", y = "Number of Monthly Trips")
```

#### Yearly
More than half of the customers of Cyclistic are members.
The only exception is the Covid-19 period but the number of members is still higher than casual riders.
```{r, echo=FALSE}
all_trips_v6 %>% 
  group_by(member_casual, year) %>% 
  summarise(number_of_rides = n()) %>%  
  arrange(member_casual, year)  %>% 
  ggplot(aes(x = year, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy",
                               "maroon")) +
  labs(title = "Yearly Number of Trips Taken", fill = "Member or Casual", x = "", y = "Number of Yearly Trips")
```

### Analysis on trip duration

#### Daily
Members are using the bikes with a similar frequency during the week.
Casual riders are preferring the weekends.
In total, weekends are the most used time of the week.
```{r, echo=FALSE}
all_trips_v6 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(ride_length = sum(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy",
                               "maroon")) +
  labs(title = "Daily Total Trip Duration", fill = "Member or Casual", x = "", y = "Daily Trip Duraiton")
```

#### Monthly
Summer has the highest demand for bikes for both customer groups.
Casual riders have a special interest during the summer.
```{r, echo=FALSE}
all_trips_v6 %>% 
  group_by(member_casual, month) %>% 
  summarize(ride_length = sum(ride_length)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy", "maroon")) +
  labs(title = "Monthly Total Trip Duration", fill = "Member or Casual", x = "", y = "Monthly Trip Duraiton")
```

#### Yearly
Until Covid-19, both customer groups were contributing similarly.
After 2019, trips taken by members decreased.
Total trip duration stayed stable because the increase in casual riders during the period compensated for the decrease in members.
```{r, echo=FALSE}
all_trips_v6%>% 
  group_by(member_casual, year) %>% 
  summarize(ride_length = sum(ride_length)) %>% 
  arrange(member_casual, year)  %>% 
  ggplot(aes(x = year, y = ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("navy",
                               "maroon")) +
  labs(title = "Yearly Total Trip Duration", fill = "Member or Casual", x = "", y = "Yearly Trip Duraiton")
``` 

### Summary of analysis
To sum up, two customer groups were contributing the revenue equally until 2019. Since then, the revenue contribution of the casual riders are higher than members even though the number of rides for casual riders is lower than for the members. This is coming from the high average trip duration of casual riders.

### Recommendations
1. Casual riders have a higher average trip duration, therefore, increasing the number of casual riders will have a more significant marginal effect than the members. Preparing special campaigns targeting casual riders will be more profitable for the company.
2. Summer has the highest demand, prepare campaigns for summer focusing on both customer groups in order to increase the number of total users instead of only focusing on the number of members. Because this attempt may cause a decrease in the number of casual riders which has a higher marginal effect on the revenue.
3. Prepare campaigns for weekends since the demand is at the top on weekends.

### Food for tought
Data includes observations about station locations of each trip. It is possible to run further analysis on the density of the stations in order to replace advertisement tools and balance the number of bikes docked at each station. Moreover, it is possible to manage the bike distribution among stations seasonally since there are data available.